Presheaves, Representables and\\the Yoneda Lemma

### 2017-07-28 notes/notes-2/notes-2.tex
Do some minor clean-ups to the notes for talk 2

\documentclass{article}

\usepackage[utf8]{inputenc}
\usepackage{amsthm}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

% \beamertemplatenavigationsymbolsempty
% \setbeamertemplate{footline}{
% 	\raisebox{0.5em}{\makebox[\paperwidth-0.5em][r]{
% 		\color{gray}\scriptsize\insertframenumber}}}

\usepackage{mathtools}
\usepackage{tikz}
\usetikzlibrary{cd, shapes, tikzmark}
\def\pathToRoot{../../}\input{\pathToRoot headers/sharedHeader}

\title{Presheaves, Representables and\\the Yoneda Lemma}
\author{Dominik Wagner}
%\date{June 7, 2017}

\begin{document}

\begin{definition}
  Let $\cat A$ and $\cat B$ be categories. A \emph{functor} $\from \cat A\to\cat B$ consists of
    \begin{itemize}
    \item a function
      \begin{align*}
        \ob(\cat A)\to\ob(\cat B),
      \end{align*}
      written as $A\mapsto F(A)$ and
    \item for each $A,A'\in\ob(\cat A)$, a function
      \begin{align*}
        \cat A(A,A')\to\cat B(F(A),F(A')),
      \end{align*}
      written as $f\mapsto F(f)$
    \end{itemize}
    such that the following axioms are satisfies:
    \begin{itemize}
    \item $F(1_A)=1_{F(A)}$ whenever $A\in\cat A$ and
    \item $F(f'\of f)=F(f')\of F(f)$ whenever $A\xrightarrow{f}A'\xrightarrow{f'}A''$ in $\cat A$.
    \end{itemize}
\end{definition}

Next, we consider a further example of a $\Set$-valued functor, which unlike the previous ones is not forgetful.
\begin{example}
  \label{ex:Mset}
  Let $\cat M$ be a category with a single object $M$, i.e. a monoid. Then, a functor $F\from\cat M\to\Set$ is a set $S=F(M)$ together with functions $F(m)\from S\to S$ for $m\from M\to M$ in $\cat M$. Hence, we can introduce the short-hand notation
  \begin{align*}
    m\cdot s=F(m)(s).
  \end{align*}
  The functorality axioms for $F$ state that
  \begin{align*}
    1_M\cdot s&=F(1_M)(s)=1_S(s)=s\\
    (m\of m')\cdot s&=F(m\of m')(s)=(F(m)\of F(m'))(s)=F(m)(F(m')(s))=m\cdot (m'\cdot s)
  \end{align*}
  for all $m,m'\from M\to M$ and $s\in S$.

  As a consequence, a functor from a one-object category, which correspond to a monoid $M$, to $\Set$ is essentially a \emph{left $M$-set}. (Recall that a left $M$-set for a monoid $M$ is a set $S$ together with a function
  \begin{align*}
    \cdot\from M\times S&\to S\\
    (m,s)&\mapsto m\cdot s
  \end{align*}
  such that $(m\of m')\cdot s=m\cdot (m'\cdot s)$ and $1_M\cdot s=s$ for all $m,m'\in M$ and $s\in S$.)  
\end{example}

@@

A very natural question to ask is whether there is also a general category theoretic analogue to monotonically decreasing functions, i.e. functions $F\from\ob(\cat P)\to\ob(\cat Q)$ between preorders $\cat P$ and $\cat Q$ such that
\begin{align*}
  P\leq_\cat P P'\rightarrow F(P)\geq_\cat Q F(P')
\end{align*}
for each $P,P'\in\ob(\cat P)$. Again, this is exactly the same as to demand that for each $P,P'\in\ob(\cat P)$ there is a function
\begin{align*}
  \cat P(P,P')\to\cat Q(F(P'),F(P)).
\end{align*}
Since there is the unique arrow in $\cat P(P,P')$ if and only if $\cat P^\op(P',P)$ contains the unique arrow we see that this amounts to the same as having a function
\begin{align*}
  \cat P^\op(P',P)\to\cat Q(F(P'),F(P)),
\end{align*}
that is we need to reverse the direction of the arrows w.r.t. $\cat P$. Due to the fact that the additional functorality axioms are trivially satisfied for preorders, we see that monotone decreasing functors are exactly the same as functors
\begin{align*}
  F\from\cat P^\op\to\cat Q
\end{align*}
for the corresponding categories $\cat P$ and $\cat Q$.

Such ``operations'' that are essentially functors but reverse the direction of the arrows are going to play a prominent role in later chapters. Therefore, they have a special name:
\begin{definition}
  Let $\cat A$ and $\cat B$ be categories. A \emph{contravariant functor} from $\cat A$ to $\cat B$ is a functor $\cat A^\op\to\cat B$.
\end{definition}
\section{Dual Vector Spaces and Contravariant Representable Functors}
Next, we are going to look at the generalization of a construction from linear algebra, which is used in the study of bilinear mappings amongst others and which can be regarded as a special contravariant functor. The general idea is that we analyze objects by analyzing the mappings into them.

@@

Bearing this in mind, the defintion of the functor $H_B$ for $B\in\cat A$ is a straight-forward generalization:
\begin{align*}
  H_B(A)=\cat A(A,B)
\end{align*}
for objects $A\in\cat A$ and for $f\from A'\to A$ the function $H_B(f)\from H_B(A)\to H_B(A')$ is defined by
\begin{align*}
  H_B(f)\from H_B(A)=\cat A(A,B)&\to H_B(A')=\cat A(A',B)\\
  g&\mapsto g\of f.
\end{align*}
\vspace{-0.7cm}
\begin{align*}
  \xymatrix{A' \ar[r]^{f} & A \ar[r]^{g} & B }
\end{align*}
The functor $H_A$ is called the \emph{contravariant representable functor} of $A$. Furthermore, contravariant functors to $\Set$ (like $H_A$) have a special name:
\begin{definition}
  Let $\cat A$ be a category. A \emph{presheaf} on $\cat A$ is a functor $\cat A^\op\to\Set$.
\end{definition}
In a later chapter we are going to discover that contravariant representable functors really help us understand the structure of locally small categories. In particular, we are going to prove that two objects $A$ and $B$ are isomorphic if and only if $H_A$ and $H_B$ are isomorphic\footnote{The notion of mappings between functors is defined in the next chapter.}, i.e. $A$ and $B$ ``look the same from all other objects''.

\begin{example}
  Similarly, as in example \ref{ex:Mset} presheaves on one-object categories (i.e. monoids $M$) correspond to \emph{right} $M$-sets, where a right $M$-set is a set $S$ together with a function
  \begin{align*}
    \cdot\from S\times M&\to S\\
    (s,m)&\mapsto s\cdot m
  \end{align*}
  such that $s\cdot (m\of m')=(s\cdot m)\cdot m'$ and $s\cdot 1_M=s$ for all $m,m'\in M$ and $s\in S$.
\end{example}

>>>>

\documentclass{article}
\def\pathToRoot{../../}\input{\pathToRoot headers/sharedHeader}

\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\usetikzlibrary{cd}

\begin{document}
\title{Presheaves, Representables and\\the Yoneda Lemma}
\author{Dominik Wagner, Felix Rech}
\maketitle

\begin{definition}
  Let $\cat A$ and $\cat B$ be categories. A \demph{functor} $F \from \cat A\to\cat B$ consists of
    \begin{itemize}
    \item a function
      \begin{align*}
        \ob(\cat A)\to\ob(\cat B),
      \end{align*}
      written as $A\mapsto F(A)$ and
    \item for each $A,A'\in\ob(\cat A)$, a function
      \begin{align*}
        \cat A(A,A')\to\cat B(F(A),F(A')),
      \end{align*}
      written as $f\mapsto F(f)$
    \end{itemize}
    such that the following axioms are satisfies:
    \begin{itemize}
    \item $F(1_A)=1_{F(A)}$ for all objects $A$ and
    \item $F(f'\of f)=F(f')\of F(f)$ for all morphisms $A\xrightarrow{f}A'\xrightarrow{f'}A''$.
    \end{itemize}
\end{definition}

Next, we consider a further example of a $\Set$-valued functor, which unlike the previous ones is not forgetful.
\begin{example}
  \label{ex:Mset}
  Let $\cat M$ be a category with a single object $M$, i.e.\ a monoid. Then, a functor $F\from\cat M\to\Set$ is a set $S=F(M)$ together with functions $F(m)\from S\to S$ for $m\from M\to M$ in $\cat M$. Hence, we can introduce the short-hand notation
  \begin{align*}
    m\cdot s=F(m)(s).
  \end{align*}
  The functorality axioms for $F$ state that
  \begin{align*}
    1_M\cdot s&=F(1_M)(s)=1_S(s)=s\\
    (m\of m')\cdot s&=F(m\of m')(s)=(F(m)\of F(m'))(s)=F(m)(F(m')(s))=m\cdot (m'\cdot s)
  \end{align*}
  for all $m,m'\from M\to M$ and $s\in S$.

  As a consequence, a functor from a one-object category, which correspond to a monoid $M$, to $\Set$ is essentially a \emph{left $M$-set}. (Recall that a left $M$-set for a monoid $M$ is a set $S$ together with a function
  \begin{align*}
    {} \cdot {} \from M\times S&\to S\\
    (m,s)&\mapsto m\cdot s
  \end{align*}
  such that $(m\of m')\cdot s=m\cdot (m'\cdot s)$ and $1_M\cdot s=s$ for all $m,m'\in M$ and $s\in S$.)
\end{example}

@@

A very natural question to ask is whether there is also a general category theoretic analogue to monotonically decreasing functions, i.e.\ functions $F\from\ob(\cat P)\to\ob(\cat Q)$ between preorders $\cat P$ and $\cat Q$ such that
\begin{align*}
  P\leq_\cat P P'\rightarrow F(P)\geq_\cat Q F(P')
\end{align*}
for each $P,P'\in\ob(\cat P)$. Again, this is exactly the same as to demand that for each $P,P'\in\ob(\cat P)$ there is a function
\begin{align*}
  \cat P(P,P')\to\cat Q(F(P'),F(P)).
\end{align*}
Since there is the unique arrow in $\cat P(P,P')$ if and only if $\cat P^\op(P',P)$ contains the unique arrow we see that this amounts to the same as having a function
\begin{align*}
  \cat P^\op(P',P)\to\cat Q(F(P'),F(P)),
\end{align*}
that is we need to reverse the direction of the arrows w.r.t. $\cat P$. Due to the fact that the additional functorality axioms are trivially satisfied for preorders, we see that monotone decreasing functors are exactly the same as functors
\begin{align*}
  F\from\cat P^\op\to\cat Q
\end{align*}
for the corresponding categories $\cat P$ and $\cat Q$.

Such ``operations'' that are essentially functors but reverse the direction of the arrows are going to play a prominent role in later chapters. Therefore, they have a special name:
\begin{definition}
  Let $\cat A$ and $\cat B$ be categories. A \demph{contravariant functor} from $\cat A$ to $\cat B$ is a functor $\cat A^\op\to\cat B$.
\end{definition}
\section{Dual Vector Spaces and Contravariant Representable Functors}
Next, we are going to look at the generalization of a construction from linear algebra which is used in the study of bilinear mappings amongst others and which can be regarded as a special contravariant functor. The general idea is that we analyze objects by analyzing the mappings into them.

@@

Bearing this in mind, the defintion of the functor $H_B$ for $B\in\cat A$ is a straight-forward generalization:
\begin{align*}
  H_B(A)=\cat A(A,B)
\end{align*}
for objects $A\in\cat A$ and for $f\from A'\to A$ the function $H_B(f)\from H_B(A)\to H_B(A')$ is defined by
\begin{align*}
  H_B(f)\from H_B(A)=\cat A(A,B)&\to H_B(A')=\cat A(A',B)\\
  g&\mapsto g\of f.
\end{align*}
\vspace{-0.7cm}
\begin{align*}
  \xymatrix{A' \ar[r]^{f} & A \ar[r]^{g} & B }
\end{align*}
The functor $H_A$ is called the \emph{contravariant representable functor} of $A$. Furthermore, contravariant functors to $\Set$ (like $H_A$) have a special name:
\begin{definition}
  Let $\cat A$ be a category. A \demph{presheaf} on $\cat A$ is a functor $\cat A^\op\to\Set$.
\end{definition}
In a later chapter we are going to discover that contravariant representable functors really help us understand the structure of locally small categories. In particular, we are going to prove that two objects $A$ and $B$ are isomorphic if and only if $H_A$ and $H_B$ are isomorphic\footnote{The notion of mappings between functors is defined in the next chapter.}, i.e. $A$ and $B$ ``look the same from all other objects''.

\begin{example}
  Similarly, as in example \ref{ex:Mset} presheaves on one-object categories (i.e.\ monoids $M$) correspond to \emph{right} $M$-sets, where a right $M$-set is a set $S$ together with a function
  \begin{align*}
    {} \cdot {} \from S\times M&\to S\\
    (s,m)&\mapsto s\cdot m
  \end{align*}
  such that $s\cdot (m\of m')=(s\cdot m)\cdot m'$ and $s\cdot 1_M=s$ for all $m,m'\in M$ and $s\in S$.
\end{example}
